### 2014-06-28 patchy-selection-paper.tex
rearranged appendices; began discussion

Expected transit time \label{eqn:mean_tau}
Expected haplotype length \label{eqn:haplotype_length}

@@

The rate at which mutations arise and colonize a patch of area $A$ is then
\begin{align} \label{eqn:mutrate}
  \mutrate % = \rho \mu \int p(x) dx  
  \approx 2 s_p \rho A \mu / \xi^2,
\end{align}
i.e.\ the product of mutational influx in the patch ($A\rho\mu$) and the probability of establishment ($2s_p/\xi^2$).
If this rate is low, then the time (in generations) until a mutation arises that
will become locally established within the patch is exponentially distributed with mean $1/\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the time scale on which new patches become colonized by the $B$ allele from new mutation is therefore approximately $1/\mutrate$.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(so, ignoring the possibility that a $B$ allele arises independently by mutation in the first patch).
To determine this, we dissect the process by which an allele transits between the patches,
in the process obtaining other useful information about the genealogy of B alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions (see Appendix \ref{apx:asymptotics}): 
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
(In one dimension, the equation can be solved to give $q(x)$ in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.)
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see Appendix \ref{apx:patchy_sims}).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  (The uptick at the right end is due to boundary effects near the edge of the range.)
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid lines shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 20 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
 \gc{The approximation and simulations disagree at long distances from
   the patch, but this is due to edge effects in the simulations.} }   \label{fig:sim_occupation_freqs}
\end{figure}

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
or in other words, the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Figure~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
To compute the rate of establishment by such migration,
we argue that these families (suitably defined) can be approximated by
independent branching processes whose spatial motion is described by Brownian motion.
Properties of these, combined with equation \eqref{eqn:eqfreq},
leads us to the expression \eqref{eqn:migrate} for the rate of effective migration,
an approximation that is valid if the distance $R$ is large (at least five times $\sigma/\sqrt{2s_m}$, say).

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles far from the patch where it is established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $\exp(-s_m)\approx 1-s_m$,
and the variance is $\xi^2$.
\plr{I changed this since equation \eqref{eqn:eqfreq} is in the small-$s_m$ limit, 
and therefore doesn't differentiate between $s_m$ and $\log(1/(1-s_m))$;
doing it this way means we don't have to define another parameter.}
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

First consider equation \eqref{eqn:gestalt_q}.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
defined so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
Concretely, it is the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x,y) dy .
\end{align}
\gc{do we need to talk about u a little more it }
\plr{not sure what to say}
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-s_m t} p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-s_m t} p_t(x,y) dt dx .
\end{align}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics{sim-transit.pdf}
  \end{center}
  \caption{
  BLAH BLAH WIGGLY LINE
  \plr{can we please keep that caption? ok, fine\dots}
  \textbf{Adaptation by migration:}
  An initially unadapted patch (10 demes at the left end of the range) is colonized around generation 5000;
  a representative lineage from the new patch is shown transiting between the two and coalescing with a lineage
  in the unadapted patch.
  In the left panel, as in Figure~\ref{fig:sim_snapshots}, color and contours show allele frequency across space and time;
  other simulation parameters are as in Figures~\ref{fig:sim_occupation_freqs}B and \ref{fig:sim_snapshots}.
  The right panel shows (in red) the length of the shared haplotype between the two lineages
  around the selected locus,
  which decreases as time progresses.
  \label{fig:lineagesmotion}
  }
\end{figure}

@@

\section{Discussion}

\plr{Note that optimal $s_m$ for migration makes migration harder to spot.}

@@

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-selection-plots.R
        \includegraphics{prob-establishment}
    \end{center}
    \label{fig:prob_estab_calcs}
    \caption{The probability of establishment of a new mutant allele as a function of distance from the edge of the region where it is beneficial,
    in an abrupt transition (left) and a gradual transition (right).
    The allele is deleterious to the left and beneficial to the right (with selection coefficients $\mp .05$ respecively);
    and the selection coefficient interpolates linearly between these in the central shaded region.
    The number of offspring is Poisson.
    The underlying landscape has demes every 15km and a migration rate of 0.5 between neighboring demes;
    the probability was found by numerically solving the equation for the probability of establishment of a multi-type branching process.
    }
\end{figure}

>>>>

Expected transit time \eqref{eqn:mean_tau}
Expected haplotype length \eqref{eqn:haplotype_length}

@@

The rate at which mutations arise and colonize a patch of area $A$ is then
\begin{align} \label{eqn:mutrate}
  \mutrate % = \rho \mu \int p(x) dx  
  \approx 2 s_p \rho A \mu / \xi^2,
\end{align}
i.e.\ the product of mutational influx in the patch ($\rho A\mu$) and the probability of establishment ($2s_p/\xi^2$).
If this rate is low, then the time (in generations) until a mutation arises that
will become locally established within the patch is exponentially distributed with mean $1/\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the time scale on which new patches become colonized by the $B$ allele from new mutation is therefore approximately $1/\mutrate$.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, we dissect the process by which an allele transits between the patches,
in the process obtaining other useful information about the genealogy of B alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions (especially regarding scaling) implicit here,
we provide a derivation in Appendix~\ref{apx:eqfreq}; 
as well as justification for the asymptotics in Appendix \ref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see below).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  (The disagreement at long distances is due to boundary effects near the edge of the range.)
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid lines shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 20 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
   }   \label{fig:sim_occupation_freqs}
\end{figure}

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
or in other words, the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Figure~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
To compute the rate of establishment by such migration,
we argue that these families (suitably defined) can be approximated by
independent branching processes whose spatial motion is described by Brownian motion.
Properties of these, combined with equation \eqref{eqn:eqfreq},
will lead us eventually to the expression \eqref{eqn:migrate} for the rate of effective migration,
an approximation that is valid if the distance $R$ is large (at least five times $\sigma/\sqrt{2s_m}$, say).

@@

To understand the process of adaptation by migration,
we therefore need to look more closely at the rare families of $B$ alleles far from the patch where it is established.
Consider one such family,
that lives far from the original patch.
As noted above, we will model this by assuming that each individual
migrates, then reproduces, independently of all other family members.
The mean squared migration distance is $\sigma^2$, takes the migrant in a random direction,
and if migration takes the individual into a patch where the $B$ allele is favored, 
we drop the individual from our family.
Since the $B$ allele is uniformly deleterious elsewhere,
we can decouple reproduction and migration
by first forming a tree in which each individual has a random number of offspring
then recursively assigning spatial locations to each individual in the tree,
and finally pruning any branches that happen to wander into a patch.
By our definition of fitness, the mean number of offspring per individual is $\exp(-s_m)\approx 1-s_m$,
and the variance is $\xi^2$.
This pruning removes both branches leading back into the already established patch (which are irrelevant)
and branches leading into the new patch (we will need to count these).
Since we are concerned with the rare migrant families that transit quickly between patches, 
back migrants to the already established patch will contribute little, since they are going the wrong direction;
so to simplify the analysis we ignore this detail.  % but could add this detail in an appendix

@@

First consider equation \eqref{eqn:gestalt_q}.
Suppose that $Z$ is one such spatial branching process as above, started at time 0 with a single individual at $x$,
defined so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the spatial region $A$.
The \emph{mean occupation density} of $Z$ can be thought of 
as the expected total number of offspring of a family beginning at $x$ that ever live at $y$.
Concretely, it is the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x,y) dy .
\end{align}
Said another way, $u(x,y)$ is the expected number of offspring of a progenitor at $x$ that will ever be observed near $y$.
Since reproduction and spatial motion are independent,
$u(x,y)$ is the sum across all generations of the expected number of individuals alive in that generation,
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-s_m t} p_t(x,y) dt .
\end{align}
The equilibrium frequency as decomposed in expression \eqref{eqn:gestalt_q} 
is the sum of mean occupation densities of a constant outflux of branching processes
from $\partial B(r_0)$:
\begin{align} \label{eqn:occupation_integral}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  . % \\
         % &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-s_m t} p_t(x,y) dt dx .
\end{align}

@@

\paragraph{Island model}
The resulting discrete process is quite general,
but if we oversimplify back to the island model, it is tractable:
suppose the probability of adaptaton by mutation is the same for each patch
and by migration is the same for each pair of patches.
The resulting process is a continuous-time version of the ``Chinese restaurant process''
described in \citet{aldous1985exchangeability} and \citet{pitman1995partitions},
and so the final partition of types across demes has the Ewens sampling distribution with parameter $\mu/m$.

Now we can compute most properties we might want about the process.
For instance, the proportion of demes that shares the same origin as a randomly sampled deme
is approximately Beta distributed -- see \cite{donnelly1989continuity} and \cite{perman1992sizebiased}.
The high connectedness of the discrete deme island model means that the expected number of distinct alleles
grows with the $\log$ of the number of demes, 
which strongly contrasts with the continuous spatial model 
where the local nature of dispersal means that doubling the species range will double the number of mutations expected. 
Furthermore, while in the continuous model the areas occupied by distinct mutations are similar in area, 
in this strong-selection discrete island model, 
only a few types tend to dominate even as the number of demes (and thus, distinct types) grows large.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{sim-transit.pdf}
  \end{center}
  \caption{
  \textbf{Adaptation by migration:}
  An initially unadapted patch (10 demes at the left end of the range) is colonized around generation 5000;
  a representative lineage from the new patch is shown transiting between the two and coalescing with a lineage
  in the unadapted patch.
  In the left panel, as in Figure~\ref{fig:sim_snapshots}, color and contours show allele frequency across space and time;
  other simulation parameters are as in Figures~\ref{fig:sim_occupation_freqs}B and \ref{fig:sim_snapshots}.
  The right panel shows (in red) the length of the shared haplotype between the two lineages
  around the selected locus,
  which decreases as time progresses.
  \label{fig:lineagesmotion}
  }
\end{figure}

@@

%%%%%%% %%%%%%% %%%%%%%
\section{Discussion} 
\label{ss:discussion}

This paper is an investigation into the basic, ill-formed question: 
What is the spatial resolution of adaptation?
In other words, what scale of environmental patchiness can a species develop and sustain local adaptations,
and over what spatial scale will evolution develop independent solutions to evolutionary problems?
The answer to this depends most strongly on $\sigma/\sqrt{s_m}$, 
the dispersal distance divided by the square root of the selection against the new allele between patches,
and much more weakly on the selective benefit within the patches or (perhaps suprisingly) the population density.
Weaker selection against alleles will make sharing of adaptations between patches easier,
but will also make it harder to spot,
since the lucky, transiting alleles move slower and carry a shorter haplotype.

Observations of similar adaptations occuring within the same species in widely separated patches
is strongly suggestive of parallel, independent adaptations.
Such observations are very interesting 
as they tell us about the degree of constraint faced by evolution --
how many different methods are there typically available to a species
with which to solve an evolutionary problem?

It is already well-appreciated that larger populations can lead to more parallelism \citep{softsweepsI}
\plr{fisher probably said this too?}
and that spatial structure contributes to this \citep{ralph2010parallel}.
Naturally, then, local adaptation to well-separated patches of some environment is even more likely to occur in parallel;
we have studied the case where patches are separated by long distances.
This is complementary to the work of \citet{XXX}, who studied local adaptation in rapidly varying environments.
Intermediate cases where habitat types are distributed in some more complex geometry
-- archipelagos of patches of varying sizes, or connected by long, skinny corridors, for instance --
remains to be studied.

To complement this theory, we need more empirical work,
in which the parameters (migration rate, sensitivity to environmental factors, etc)
are reasonably well-understood.
However, learning about these aspects is difficult in the field,
and it is still not understood how much of the intraspecific variation we observe
is due to adaptation to local conditions, or even how to tell.
Notable steps forward include \citet{turner2010serpentine}, Mimulus, Amity, others?

\paragraph{Standing variation} 
We have tended to focus on relative rates of adaptation,
since we are thinking of applications in which we know that adaptation has occurred,
and the question is whether or not adaptations in distinct patches
have adapted indpendently or not.
However, especially if rates of adaptation by new mutation are very low,
any adaptation that does occur may have to make use of standing variation.
The unstructured case was studied by \citet{softsweepsI}
and we study the continuous, spatial case in \citet{toappear}.
If local, parallel adaptation is from standing variation,
then the dynamics of adaptation should not depend strongly on migration patterns
(but note that the spatial distribution of standing variation will be).

\paragraph{Dominance}
We have mostly swept the issue of dominance under the rug
by dealing with essentially haploid models,
and appealing to the fact that the dynamics we study occur where the mutation is rare,
and hence mostly present only in heterozygotes.
If, however, the mutation is recessive, then it is essentially neutral where rare,
and so could have a much easier time spreading between patches.
This is counteracted, however, by the increased difficulty with which the mutation would establish
once arriving in a new patch.
On the other hand, a dominant mutation would not differ greatly from the theory presented here.
In fact, MC1r is ?dominant? and agouti is ?recessive?
Other schemes could be obtained.

optimal $s_m$ for migration makes migration harder to spot

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-selection-plots.R
        \includegraphics{prob-establishment}
    \end{center}
    \caption{The probability of establishment of a new mutant allele as a function of distance from the edge of the region where it is beneficial,
    in an abrupt transition (left) and a gradual transition (right).
    The allele is deleterious to the left and beneficial to the right (with selection coefficients $\mp 0.05$ respecively);
    and the selection coefficient interpolates linearly between these in the central shaded region.
    The number of offspring is Poisson.
    The underlying landscape has demes every 15km and a migration rate of 0.5 between neighboring demes;
    the probability was found by numerically solving the equation for the probability of establishment of a multi-type branching process.
    \label{fig:prob_estab_calcs}
    }
\end{figure}

@@

\section{UNUSED AFTER HERE?}
